---
title: "Assignment 1: Census Data Quality for Policy Decisions"
subtitle: "Evaluating Data Reliability for Algorithmic Decision-Making"
author: "Zhiyuan Zhao"
date: 2025/09/18
format: 
  html:
    code-fold: false
    toc: true
    toc-location: left
    theme: cosmo
execute:
  warning: false
  message: false
---

# Assignment Overview

## Scenario

You are a data analyst for the **[Your State] Department of Human Services**. The department is considering implementing an algorithmic system to identify communities that should receive priority for social service funding and outreach programs. Your supervisor has asked you to evaluate the quality and reliability of available census data to inform this decision.

Drawing on our Week 2 discussion of algorithmic bias, you need to assess not just what the data shows, but how reliable it is and what communities might be affected by data quality issues.

## Learning Objectives

- Apply dplyr functions to real census data for policy analysis
- Evaluate data quality using margins of error 
- Connect technical analysis to algorithmic decision-making
- Identify potential equity implications of data reliability issues
- Create professional documentation for policy stakeholders

## Submission Instructions

**Submit by posting your updated portfolio link on Canvas.** Your assignment should be accessible at `your-portfolio-url/assignments/assignment_1/`

Make sure to update your `_quarto.yml` navigation to include this assignment under an "Assignments" menu.

# Part 1: Portfolio Integration

Create this assignment in your portfolio repository under an `assignments/assignment_1/` folder structure. Update your navigation menu to include:

```
- text: Assignments
  menu:
    - href: assignments/assignment_1/your_file_name.qmd
      text: "Assignment 1: Census Data Exploration"
```
If there is a special character like comma, you need use double quote mark so that the quarto can identify this as text

# Setup

```{r setup}

# Load required packages (hint: you need tidycensus, tidyverse, and knitr)
library(tidycensus)
library(tidyverse)
library(knitr)
# Set your Census API key
census_api_key("a8081579173d9c4b9d21b9265f9c67c8708d6a02")
# Choose your state for analysis - assign it to a variable called my_state
my_state <- "Pennsylvania"

```

**State Selection:** I have chosen Pennsylvnia for this analysis because: Now I live in philadelpha.

# Part 2: County-Level Resource Assessment

## 2.1 Data Retrieval

**Your Task:** Use `get_acs()` to retrieve county-level data for your chosen state.

**Requirements:**
- Geography: county level
- Variables: median household income (B19013_001) and total population (B01003_001)  
- Year: 2022
- Survey: acs5
- Output format: wide

**Hint:** Remember to give your variables descriptive names using the `variables = c(name = "code")` syntax.

```{r county-data}
# Write your get_acs() code here
state_pop <- get_acs(
  geography = "state",
  variables = "B01003_001",  # Total population
  year = 2022,
  survey = "acs5"
)
median_income  <- get_acs(
  geography = "state",
  variables = "B19013_001",  # Total population
  year = 2022,
  survey = "acs5"
)

glimpse(state_pop)
pa_data <- get_acs(
  geography = "county",
  variables = c(
    total_pop = "B01003_001",
    median_income = "B19013_001"
  ),
  state = "PA",
  year = 2022,
  output = "wide"  # Makes analysis easier
)

head(pa_data)
# Clean the county names to remove state name and "County" 

# Hint: use mutate() with str_remove()
pa_clean <- pa_data %>%
  mutate(
    # Remove state name from county names
    county_name = str_remove(NAME, ", Pennsylvania"),
    # Remove "County" word
    county_name = str_remove(county_name, " County")
  )
# Display the first few rows
head(pa_clean)
```

## 2.2 Data Quality Assessment

**Your Task:** Calculate margin of error percentages and create reliability categories.

**Requirements:**
- Calculate MOE percentage: (margin of error / estimate) * 100
- Create reliability categories:
  - High Confidence: MOE < 5%
  - Moderate Confidence: MOE 5-10%  
  - Low Confidence: MOE > 10%
- Create a flag for unreliable estimates (MOE > 10%)

**Hint:** Use `mutate()` with `case_when()` for the categories.

```{r income-reliability}
# Calculate MOE percentage and reliability categories using mutate()
select(pa_clean, NAME, county_name)

pa_reliability <- pa_clean %>%
  mutate(
    # Calculate MOE as percentage of estimate
    moe_percentage = round((median_incomeM / median_incomeE) * 100, 2),
    
    # Create reliability categories
    reliability = case_when(
      moe_percentage < 5 ~ "High Confidence",
      moe_percentage >= 5 & moe_percentage <= 10 ~ "Moderate",
      moe_percentage > 10 ~ "Low Confidence"
    )
  )


# Create a summary showing count of counties in each reliability category
# Hint: use count() and mutate() to add percentages
pa_reliability_summary <- pa_reliability %>%
  count(reliability) %>%
  mutate(percent = n / sum(n) * 100)

pa_reliability_summary

```

## 2.3 High Uncertainty Counties

**Your Task:** Identify the 5 counties with the highest MOE percentages.

**Requirements:**
- Sort by MOE percentage (highest first)
- Select the top 5 counties
- Display: county name, median income, margin of error, MOE percentage, reliability category
- Format as a professional table using `kable()`

**Hint:** Use `arrange()`, `slice()`, and `select()` functions.

```{r high-uncertainty}
top5_uncertain <- pa_reliability %>%
  arrange(desc(moe_percentage)) %>%
  slice(1:5) %>%
  select(
    County = county_name,
    `Median Income` = median_incomeE,
    `Margin of Error` = median_incomeM,
    `MOE (%)` = moe_percentage,
    Reliability = reliability
  )

kable(
  top5_uncertain,
  digits = 2,
  caption = "Top 5 Pennsylvania Counties with Highest Income Estimate Uncertainty"
)
```

**Data Quality Commentary:**

These results show that some counties have considerable uncertainty in their median income estimates. For algorithmic decision-making, relying directly on these values could produce biased resource allocation. Counties with higher MOE may be systematically underestimated or overestimated, which could affect their ability to receive funding or services. Factors contributing to higher uncertainty likely include small population sizes, greater variability in income distributions, or difficulties in collecting survey responses in certain regions.

# Part 3: Neighborhood-Level Analysis

## 3.1 Focus Area Selection

**Your Task:** Select 2-3 counties from your reliability analysis for detailed tract-level study.

**Strategy:** Choose counties that represent different reliability levels (e.g., 1 high confidence, 1 moderate, 1 low confidence) to compare how data quality varies.

```{r select-counties}
# Use filter() to select 2-3 counties from your county_reliability data
# Store the selected counties in a variable called selected_counties
selected_counties <- pa_reliability %>%
 filter(county_name %in% c("Allegheny", "Elk")) %>%
  select(county_name, median_incomeE, reliability, moe_percentage)


# Display the selected counties with their key characteristics
# Show: county name, median income, MOE percentage, reliability category
selected_counties
```

**Comment on the output:** [write something :)]
Selecting Allegheny County and Elk County allows for a meaningful comparison across different levels of reliability. Allegheny County, with its large population, produces relatively stable data, while Elk County, with fewer residents and greater income variability, exhibits higher MOE. Comparing these two counties illustrates a broader pattern: larger counties tend to have more reliable data, while smaller or rural counties face greater challenges.

## 3.2 Tract-Level Demographics

**Your Task:** Get demographic data for census tracts in your selected counties.

**Requirements:**
- Geography: tract level
- Variables: white alone (B03002_003), Black/African American (B03002_004), Hispanic/Latino (B03002_012), total population (B03002_001)
- Use the same state and year as before
- Output format: wide
- **Challenge:** You'll need county codes, not names. Look at the GEOID patterns in your county data for hints.

```{r tract-demographics}
# Define your race/ethnicity variables with descriptive names
race_vars <- c(
  total_pop = "B03002_001",
  white = "B03002_003",
  black = "B03002_004",
  hispanic = "B03002_012"
)
# Use get_acs() to retrieve tract-level data
# Hint: You may need to specify county codes in the county parameter
tract_data <- get_acs(
  geography = "tract",
  state = "PA",
  county = c("003", "047"),  # Selected county GEOIDs
  variables = race_vars,
  year = 2022,
  survey = "acs5",
  output = "wide"
)
# Calculate percentage of each group using mutate()
tract_data <- tract_data %>%
  mutate(
    pct_white = round((whiteE / total_popE) * 100, 2),
    pct_black = round((blackE / total_popE) * 100, 2),
    pct_hispanic = round((hispanicE / total_popE) * 100, 2)
  )
# Create percentages for white, Black, and Hispanic populations

# Add readable tract and county name columns using str_extract() or similar
tract_data <- tract_data %>%
  mutate(
    tract_name = str_extract(NAME, "Tract [0-9\\.]+"),
    county_name = str_replace(NAME, " Tract [0-9\\.]+, Pennsylvania", "")
  )
head(tract_data)
```

## 3.3 Demographic Analysis

**Your Task:** Analyze the demographic patterns in your selected areas.

```{r demographic-analysis}
# Find the tract with the highest percentage of Hispanic/Latino residents
# Hint: use arrange() and slice() to get the top tract
top_hispanic_tract <- tract_data %>%
  arrange(desc(pct_hispanic)) %>%
  slice(1) %>%
  select(county_name, tract_name, pct_hispanic)

top_hispanic_tract
# Calculate average demographics by county using group_by() and summarize()
county_summary <- tract_data %>%
  filter(substr(GEOID, 1, 5) %in% c("42003", "42047")) %>%  
  group_by(county_geoid = substr(GEOID, 1, 5)) %>%          
  summarize(
    county_name = first(county_name),                        
    n_tracts = n(),                                        
    avg_pct_white = round(mean(pct_white, na.rm = TRUE), 2),
    avg_pct_black = round(mean(pct_black, na.rm = TRUE), 2),
    avg_pct_hispanic = round(mean(pct_hispanic, na.rm = TRUE), 2),
    .groups = "drop"
  )

# Show: number of tracts, average percentage for each racial/ethnic group
# Create a nicely formatted table of your results using kable()
kable(
  county_summary,
  caption = "Average Tract-Level Demographics by Selected County",
  digits = 2
)
```

# Part 4: Comprehensive Data Quality Evaluation

## 4.1 MOE Analysis for Demographic Variables

**Your Task:** Examine margins of error for demographic variables to see if some communities have less reliable data.

**Requirements:**
- Calculate MOE percentages for each demographic variable
- Flag tracts where any demographic variable has MOE > 15%
- Create summary statistics

```{r demographic-moe}
# Calculate MOE percentages for white, Black, and Hispanic variables
# Hint: use the same formula as before (margin/estimate * 100)
tract_data <- tract_data %>%
  mutate(
    moe_pct_white = ifelse(whiteE > 0, round((whiteM / whiteE) * 100, 2), NA),
    moe_pct_black = ifelse(blackE > 0, round((blackM / blackE) * 100, 2), NA),
    moe_pct_hispanic = ifelse(hispanicE > 0, round((hispanicM / hispanicE) * 100, 2), NA)
  )
# Create a flag for tracts with high MOE on any demographic variable
# Use logical operators (| for OR) in an ifelse() statement
tract_data <- tract_data %>%
  mutate(
    high_moe_flag = ifelse(
      moe_pct_white > 15 | moe_pct_black > 15 | moe_pct_hispanic > 15,
      1, 0
    )
  )

# Create summary statistics showing how many tracts have data quality issues
moe_summary <- tract_data %>%
  mutate(county_geoid = substr(GEOID, 1, 5)) %>%
  filter(county_geoid %in% c("42003", "42047")) %>%
  group_by(county_geoid) %>%
  summarize(
    n_tracts = n(),
    n_high_moe = sum(high_moe_flag, na.rm = TRUE),
    pct_high_moe = round((n_high_moe / n_tracts) * 100, 2),
    avg_moe_white = round(mean(moe_pct_white, na.rm = TRUE), 2),
    avg_moe_black = round(mean(moe_pct_black, na.rm = TRUE), 2),
    avg_moe_hispanic = round(mean(moe_pct_hispanic, na.rm = TRUE), 2),
    .groups = "drop"
  ) %>%
  mutate(
    county_name = case_when(
      county_geoid == "42003" ~ "Allegheny County",
      county_geoid == "42047" ~ "Elk County"
    )
  ) %>%
  select(county_name, everything())
# Nicely formatted summary table
kable(
  moe_summary,
  caption = "Summary of MOE Reliability Issues by County",
  digits = 2
)
```

## 4.2 Pattern Analysis

**Your Task:** Investigate whether data quality problems are randomly distributed or concentrated in certain types of communities.

```{r pattern-analysis}
# Group tracts by whether they have high MOE issues
# Calculate average characteristics for each group:
# - population size, demographic percentages
pattern_summary <- tract_data %>%
  filter(substr(GEOID, 1, 5) %in% c("42003", "42047")) %>%
  group_by(high_moe_flag) %>%
  summarize(
    n_tracts = n(),
    avg_total_pop    = round(mean(total_popE,    na.rm = TRUE), 0),
    avg_pct_white    = round(mean(pct_white,     na.rm = TRUE), 2),
    avg_pct_black    = round(mean(pct_black,     na.rm = TRUE), 2),
    avg_pct_hispanic = round(mean(pct_hispanic,  na.rm = TRUE), 2),
    .groups = "drop"
  ) %>%
  mutate(
    moe_group = case_when(
      high_moe_flag == 1 ~ "High MOE Tracts",
      high_moe_flag == 0 ~ "Low MOE Tracts",
      is.na(high_moe_flag) ~ "Low MOE Tracts"  # treat NA as Low MOE
    )
  ) %>%
  select(moe_group, everything(), -high_moe_flag)
# Use group_by() and summarize() to create this comparison
# Create a professional table showing the patterns
kable(
  pattern_summary,
  caption = "Comparison of Demographic Patterns by Data Quality Group",
  digits = 2
)
```

**Pattern Analysis:**
The results suggest that tracts with higher MOE tend to have smaller populations and, in some cases, higher proportions of minority residents. This indicates that data quality problems are not randomly distributed but are concentrated in communities that are either less populated or more diverse. Such communities are at risk of being misclassified or overlooked in algorithmic systems due to unreliable data, reinforcing inequities. The root causes likely include small sample sizes, challenges in achieving representative responses, and structural disparities between urban and rural areas.


# Part 5: Policy Recommendations

## 5.1 Analysis Integration and Professional Summary

**Your Task:** Write an executive summary that integrates findings from all four analyses.

**Executive Summary Requirements:**
1. **Overall Pattern Identification**: Across the state, larger counties and tracts generally demonstrate higher data reliability, while smaller or less populated areas tend to exhibit higher MOE. This pattern holds for both income and demographic data.
2. **Equity Assessment**: Communities with low data reliability face the greatest risk of algorithmic bias. Smaller counties and tracts with higher minority populations are particularly vulnerable to being disadvantaged if algorithmic systems undervalue their needs due to measurement uncertainty.
3. **Root Cause Analysis**: Underlying drivers of these data quality issues include the limitations of sampling design in sparsely populated areas, underrepresentation of minority populations in survey data, and broader geographic disparities in data collection. These structural challenges affect both the accuracy of estimates and the risk of bias when such data informs policy.
4. **Strategic Recommendations**: For algorithmic implementation, the department should rely more heavily on automated methods in high-confidence regions, apply continuous monitoring and auditing in moderate-confidence regions, and require manual review or supplemental data collection in low-confidence regions. Longer term, efforts should focus on improving representation and data collection practices in underserved communities to reduce systematic inequality in decision-making.

**Executive Summary:**

This analysis finds that larger counties like Allegheny generally produce reliable ACS estimates, while smaller counties such as Elk often face high margins of error. At the tract level, data uncertainty is more pronounced in low-population and more diverse communities, raising equity concerns if algorithms rely only on point estimates. These reliability gaps are largely driven by small sample sizes and uneven survey representation. To reduce bias, high-confidence areas can safely use algorithmic allocation, moderate-confidence areas require monitoring, and low-confidence areas should include manual review or supplemental data.

## 6.3 Specific Recommendations

**Your Task:** Create a decision framework for algorithm implementation.

```{r recommendations-data}
# Create a summary table using your county reliability data
# Include: county name, median income, MOE percentage, reliability category
recommendations_table <- pa_reliability %>%
  # Add algorithm recommendation based on reliability
  mutate(
    algorithm_recommendation = case_when(
      reliability == "High Confidence" ~ "Safe for algorithmic decisions",
      reliability == "Moderate"        ~ "Use with caution - monitor outcomes",
      reliability == "Low Confidence"  ~ "Requires manual review or additional data",
      TRUE ~ "Unknown"
    )
  ) %>%
  # Keep only the desired columns
  select(county_name, median_incomeE, moe_percentage, reliability, algorithm_recommendation)

# Add a new column with algorithm recommendations using case_when():
# - High Confidence: "Safe for algorithmic decisions"
# - Moderate Confidence: "Use with caution - monitor outcomes"  
# - Low Confidence: "Requires manual review or additional data"

# Format as a professional table with kable()
kable(
  recommendations_table,
  caption = "Algorithm Implementation Recommendations by County",
  digits = 2
)
```

**Key Recommendations:**

**Your Task:** Use your analysis results to provide specific guidance to the department.

1. **Counties suitable for immediate algorithmic implementation:** High-confidence counties, such as Allegheny County, have low error margins and can be safely included in algorithmic decision-making for efficient resource allocation.

2. **Counties requiring additional oversight:** Counties with moderate confidence can still be included in algorithmic processes, but outcomes should be regularly monitored and audited to ensure no systematic misallocation occurs.

3. **Counties needing alternative approaches:** Low-confidence counties, such as Elk County, should not rely solely on algorithms. Instead, these areas require manual review, supplemental surveys, or hybrid decision-making approaches to ensure fair distribution of resources.

## Questions for Further Investigation

1. Are high-MOE tracts geographically clustered (e.g., rural areas, urban peripheries), and how might these spatial patterns affect regional equity in resource allocation?
2. How has the reliability of ACS estimates changed over time (e.g., 2010–2022), and are some communities experiencing persistent issues with data quality that could systematically disadvantage them?

# Technical Notes

**Data Sources:** 
- U.S. Census Bureau, American Community Survey 2018-2022 5-Year Estimates
- Retrieved via tidycensus R package on September 18, 2025

**Reproducibility:** 
- All analysis conducted in R version 4.5.1
- Census API key required for replication
- Complete code and documentation available at: https://musa-5080-fall-2025.github.io/portfolio-setup-ZhiyuanZhaoMicheal/

**Methodology Notes:**
	•	County names were cleaned by removing state and “County” suffixes to improve readability.
	•	Reliability categories were defined using fixed MOE percentage thresholds (<5%, 5–10%, >10%) for income, and a 15% cutoff for demographic variables.
	•	Two counties (Allegheny and Elk) were selected for tract-level analysis to illustrate contrasting reliability levels.
	•	Demographic percentages were calculated relative to total tract population, and summary statistics were aggregated at the county leve

**Limitations:**
	•	ACS data are sample-based, so small counties and tracts often have wide MOEs that reduce precision.
	•	The analysis covers only Pennsylvania; results may not generalize to other states.
	•	Using 5-year ACS estimates improves stability but limits the ability to detect short-term changes.
	•	Data quality issues may be particularly pronounced for smaller racial/ethnic groups not included in this study.

---

## Submission Checklist

Before submitting your portfolio link on Canvas:

- [ ] All code chunks run without errors
- [ ] All "[Fill this in]" prompts have been completed
- [ ] Tables are properly formatted and readable
- [ ] Executive summary addresses all four required components
- [ ] Portfolio navigation includes this assignment
- [ ] Census API key is properly set 
- [ ] Document renders correctly to HTML

**Remember:** Submit your portfolio URL on Canvas, not the file itself. Your assignment should be accessible at `your-portfolio-url/assignments/assignment_1/your_file_name.html`