---
title: "Assignment 2: Spatial Analysis and Visualization"
subtitle: "Healthcare Access and Equity in Pennsylvania"
author: "Zhiyuan Zhao"
date: 2025/10/02
format: 
  html:
    code-fold: false
    toc: true
    toc-location: left
    theme: cosmo
    embed-resources: true
execute:
  warning: false
  message: false
---

## Assignment Overview

**Learning Objectives:** - Apply spatial operations to answer policy-relevant research questions - Integrate census demographic data with spatial analysis - Create publication-quality visualizations and maps - Work with spatial data from multiple sources - Communicate findings effectively for policy audiences

------------------------------------------------------------------------

## Part 1: Healthcare Access for Vulnerable Populations

### Research Question

**Which Pennsylvania counties have the highest proportion of vulnerable populations (elderly + low-income) living far from hospitals?**

Your analysis should identify counties that should be priorities for healthcare investment and policy intervention.

### Required Analysis Steps

Complete the following analysis, documenting each step with code and brief explanations:

#### Step 1: Data Collection (5 points)

Load the required spatial data: - Pennsylvania county boundaries - Pennsylvania hospitals (from lecture data) - Pennsylvania census tracts

**Your Task:**

```{r}
# Load required packages
library(sf)
library(tidyverse)
library(tidycensus)
library(tigris)
library(ggplot2)
library(knitr)
library(kableExtra)
library(units)
library(patchwork)
library(ggspatial)

census_api_key("a8081579173d9c4b9d21b9265f9c67c8708d6a02", install = TRUE, overwrite = TRUE)

# Load spatial data
library(here)

pa_counties <- st_read(here::here("Assignments", "assignment_2", "data", "Pennsylvania_County_Boundaries.shp"))
hospitals   <- st_read(here::here("Assignments", "assignment_2", "data", "hospitals.geojson"))
# Pennsylvania census tracts
pa_tracts <- tracts("PA", year = 2022, cb = TRUE)

# Check that all data loaded correctly
cat("Number of hospitals:", nrow(hospitals), "\n")
cat("Number of census tracts", nrow(pa_tracts), "\n")
# CRS info
cat("Counties CRS:", st_crs(pa_counties)$input, "\n")
cat("Hospitals CRS:", st_crs(hospitals)$input, "\n")
cat("Census tracts CRS:", st_crs(pa_tracts)$input, "\n")

```

**Questions to answer:** 
1. How many hospitals are in your dataset? 
223
2. How many census tracts?
3445 
3. What coordinate reference system is each dataset in?
Counties CRS: WGS 84 / Pseudo-Mercator 
Hospitals CRS: WGS 84 
Census tracts CRS: NAD83 
------------------------------------------------------------------------

#### Step 2: Get Demographic Data

Use `tidycensus` to download tract-level demographic data for Pennsylvania.

**Required variables:** - Total population - Median household income - Population 65 years and over (you may need to sum multiple age categories)

**Your Task:**

```{r}
# Get demographic data from ACS
pa_demo <- get_acs(
  geography = "tract",
  state = "PA",
  variables = c(
    total_pop = "B01003_001",           
    median_income = "B19013_001",       
    pop_65_74 = "B01001_020",          
    pop_75_84 = "B01001_021",           
    pop_85p = "B01001_022"         
  ),
  year = 2022,
  geometry = TRUE
)

# Pivot the data to wide format
pa_demo_wide <- pa_demo %>%
  select(GEOID, NAME, variable, estimate) %>%
  pivot_wider(
    names_from = variable, values_from = estimate,
    values_fill = 0
  ) %>%
  mutate(
    pop_65p = rowSums(across(c(pop_65_74, pop_75_84, pop_85p)), na.rm = TRUE),
    elderly_pct = 100 * pop_65p / total_pop
  )

# Join to tract boundaries
tracts_with_demo <- pa_demo_wide

# Check for missing data
cat("Missing income data:", sum(is.na(tracts_with_demo$median_income)), "\n")
cat("PA Median income: $", 
    median(tracts_with_demo$median_income, na.rm = TRUE), "\n")

```

**Questions to answer:** 
- What year of ACS data are you using? 
2022
- How many tracts have missing income data?
63
- What is the median income across all PA census tracts?
70188

------------------------------------------------------------------------

#### Step 3: Define Vulnerable Populations

Identify census tracts with vulnerable populations based on TWO criteria: 1. Low median household income (choose an appropriate threshold) 2. Significant elderly population (choose an appropriate threshold)

**Your Task:**

```{r}
# Filter for vulnerable tracts based on your criteria
# Define thresholds
income_threshold <- 50000   
elderly_threshold <- 10    

# Identify vulnerable tracts
vulnerable_tracts <- tracts_with_demo %>%
  filter(!is.na(median_income)) %>%
  mutate(
    low_income = median_income < income_threshold,
    sign_elderly = elderly_pct > elderly_threshold,
    vulnerable = low_income | sign_elderly
  )

# Summaries
summary_vals <- list(
  pa_median = median(tracts_with_demo$median_income, na.rm = TRUE),
  n_vul = sum(vulnerable_tracts$vulnerable),
  pct_vul = round(mean(vulnerable_tracts$vulnerable) * 100, 1),
  n_low = sum(vulnerable_tracts$low_income),
  n_old = sum(vulnerable_tracts$sign_elderly)
)

cat(glue::glue(
  "Income threshold: ${income_threshold} \n",
  "Elderly threshold: {elderly_threshold}%\n",
  "Vulnerable tracts: {summary_vals$n_vul} ({summary_vals$pct_vul}%)\n",
  "Low median household income: {summary_vals$n_low}\n",
  "Significant elderly population: {summary_vals$n_old}\n"
))



```

**Questions to answer:** 
- What income threshold did you choose and why? 
Income threshold chosen: $50,000.
This represents census tracts where the median household income is below roughly 70% of Pennsylvania’s overall median, marking areas that are relatively economically disadvantaged.

- What elderly population threshold did you choose and why? 
Elderly population threshold chosen: 10%.
Tracts where more than 10% of residents are aged 65 or older were considered to have a significant elderly population, a common demographic baseline in vulnerability analyses.

- How many tracts meet your vulnerability criteria? 
700 census tracts meet at least one of the two conditions 
- What percentage of PA census tracts are considered vulnerable by your definition?
20.7% 

------------------------------------------------------------------------

#### Step 4: Calculate Distance to Hospitals

For each vulnerable tract, calculate the distance to the nearest hospital.

**Your Task:**

```{r}
# Transform to appropriate projected CRS
# EPSG 2272 is optimized for southern Pennsylvania counties such as Philadelphia, Lancaster, Chester, and York
target_crs <- 2272

# Reproject all layers
pa_counties_p <- st_transform(pa_counties, target_crs)
hospitals_p   <- st_transform(hospitals,   target_crs)
vul_p         <- st_transform(vulnerable_tracts, target_crs)

# Calculate distance from each tract centroid to nearest hospital
# tract centroids
vul_centroids <- st_centroid(vul_p)

# nearest hospital index
nn_idx <- st_nearest_feature(vul_centroids, hospitals_p)

# distance in feet -> miles
dists_ft <- st_distance(vul_centroids, hospitals_p[nn_idx, ], by_element = TRUE)
dists_mi <- units::set_units(dists_ft, "mi") |> units::drop_units()

# attach results
vulnerable_tracts_dist <- vul_p |>
  mutate(
    nearest_hosp_idx = as.integer(nn_idx),
    dist_to_hosp_mi  = dists_mi
  )

# optional: join hospital name if a name field exists
name_field <- intersect(c("NAME","Name","name","HOSP_NAME","HOSPITAL","HospName"), names(hospitals_p))[1]
if (!is.na(name_field)) {
  vulnerable_tracts_dist <- vulnerable_tracts_dist |>
    mutate(nearest_hosp_name = hospitals_p[[name_field]][nearest_hosp_idx])
}

# summary statistics
avg_dist <- mean(vulnerable_tracts_dist$dist_to_hosp_mi, na.rm = TRUE)
min_dist <- min(vulnerable_tracts_dist$dist_to_hosp_mi, na.rm = TRUE)
max_dist <- max(vulnerable_tracts_dist$dist_to_hosp_mi, na.rm = TRUE)
n_over15 <- sum(vulnerable_tracts_dist$dist_to_hosp_mi > 15, na.rm = TRUE)

cat(
  "Average distance to nearest hospital:", round(avg_dist, 2), "miles\n",
  "Minimum distance:", round(min_dist, 2), "miles\n",
  "Maximum distance:", round(max_dist, 2), "miles\n",
  "Number of tracts more than 15 miles from nearest hospital:", n_over15, "\n"
)



```

**Requirements:** - Use an appropriate projected coordinate system for Pennsylvania - Calculate distances in miles - Explain why you chose your projection

**Questions to answer:** 
- What is the average distance to the nearest hospital for vulnerable tracts? 
4.43 miles

- What is the maximum distance? 
32.84 miles

- How many vulnerable tracts are more than 15 miles from the nearest hospital?
127

------------------------------------------------------------------------

#### Step 5: Identify Underserved Areas

Define "underserved" as vulnerable tracts that are more than 15 miles from the nearest hospital.

**Your Task:**

```{r}
# Create underserved variable
# Definition: vulnerable tracts with distance > 15 miles

vulnerable_tracts_dist <- vulnerable_tracts_dist %>%
  mutate(underserved = dist_to_hosp_mi > 15)

# Counts and percentages
n_underserved  <- sum(vulnerable_tracts_dist$underserved, na.rm = TRUE)
n_with_dist    <- sum(!is.na(vulnerable_tracts_dist$dist_to_hosp_mi))
n_vulnerable   <- nrow(vulnerable_tracts_dist)
pct_underserved <- round(100 * n_underserved / n_with_dist, 1)

cat(
  "Underserved tracts:", n_underserved, "\n",
  "Percent underserved (of tracts with a distance):", pct_underserved, "%\n",
  "Vulnerable tracts total:", n_vulnerable, "\n",
  "With valid distance:", n_with_dist, "\n"
)



```

**Questions to answer:** - How many tracts are underserved? - What percentage of vulnerable tracts are underserved? - Does this surprise you? Why or why not?

------------------------------------------------------------------------

#### Step 6: Aggregate to County Level

Use spatial joins and aggregation to calculate county-level statistics about vulnerable populations and hospital access.

**Your Task:**

```{r}
# Spatial join tracts to counties


# Aggregate statistics by county




```

**Required county-level statistics:** - Number of vulnerable tracts - Number of underserved tracts\
- Percentage of vulnerable tracts that are underserved - Average distance to nearest hospital for vulnerable tracts - Total vulnerable population

**Questions to answer:** - Which 5 counties have the highest percentage of underserved vulnerable tracts? - Which counties have the most vulnerable people living far from hospitals? - Are there any patterns in where underserved counties are located?

------------------------------------------------------------------------

#### Step 7: Create Summary Table

Create a professional table showing the top 10 priority counties for healthcare investment.

**Your Task:**

```{r}
# Create and format priority counties table




```

**Requirements:** - Use `knitr::kable()` or similar for formatting - Include descriptive column names - Format numbers appropriately (commas for population, percentages, etc.) - Add an informative caption - Sort by priority (you decide the metric)

------------------------------------------------------------------------

## Part 2: Comprehensive Visualization

Using the skills from Week 3 (Data Visualization), create publication-quality maps and charts.

### Map 1: County-Level Choropleth

Create a choropleth map showing healthcare access challenges at the county level.

**Your Task:**

```{r}
# Create county-level access map




```

**Requirements:** - Fill counties by percentage of vulnerable tracts that are underserved - Include hospital locations as points - Use an appropriate color scheme - Include clear title, subtitle, and caption - Use `theme_void()` or similar clean theme - Add a legend with formatted labels

------------------------------------------------------------------------

### Map 2: Detailed Vulnerability Map

Create a map highlighting underserved vulnerable tracts.

**Your Task:**

```{r}
# Create detailed tract-level map




```

**Requirements:** - Show underserved vulnerable tracts in a contrasting color - Include county boundaries for context - Show hospital locations - Use appropriate visual hierarchy (what should stand out?) - Include informative title and subtitle

------------------------------------------------------------------------

### Chart: Distribution Analysis

Create a visualization showing the distribution of distances to hospitals for vulnerable populations.

**Your Task:**

```{r}
# Create distribution visualization




```

**Suggested chart types:** - Histogram or density plot of distances - Box plot comparing distances across regions - Bar chart of underserved tracts by county - Scatter plot of distance vs. vulnerable population size

**Requirements:** - Clear axes labels with units - Appropriate title - Professional formatting - Brief interpretation (1-2 sentences as a caption or in text)

------------------------------------------------------------------------

## Part 3: Bring Your Own Data Analysis

Choose your own additional spatial dataset and conduct a supplementary analysis.

### Challenge Options

Choose ONE of the following challenge exercises, or propose your own research question using OpenDataPhilly data (https://opendataphilly.org/datasets/).

**Note these are just loose suggestions to spark ideas - follow or make your own as the data permits and as your ideas evolve. This analysis should include bringing in your own dataset, ensuring the projection/CRS of your layers align and are appropriate for the analysis (not lat/long or geodetic coordinate systems). The analysis portion should include some combination of spatial and attribute operations to answer a relatively straightforward question**

------------------------------------------------------------------------

#### Education & Youth Services

**Option A: Educational Desert Analysis** - **Data:** Schools, Libraries, Recreation Centers, Census tracts (child population) - **Question:** "Which neighborhoods lack adequate educational infrastructure for children?" - **Operations:** Buffer schools/libraries (0.5 mile walking distance), identify coverage gaps, overlay with child population density - **Policy relevance:** School district planning, library placement, after-school program siting

**Option B: School Safety Zones** - **Data:** Schools, Crime Incidents, Bike Network - **Question:** "Are school zones safe for walking/biking, or are they crime hotspots?" - **Operations:** Buffer schools (1000ft safety zone), spatial join with crime incidents, assess bike infrastructure coverage - **Policy relevance:** Safe Routes to School programs, crossing guard placement

------------------------------------------------------------------------

#### Environmental Justice

**Option C: Green Space Equity** - **Data:** Parks, Street Trees, Census tracts (race/income demographics) - **Question:** "Do low-income and minority neighborhoods have equitable access to green space?" - **Operations:** Buffer parks (10-minute walk = 0.5 mile), calculate tree canopy or park acreage per capita, compare by demographics - **Policy relevance:** Climate resilience, environmental justice, urban forestry investment ---

#### Public Safety & Justice

**Option D: Crime & Community Resources** - **Data:** Crime Incidents, Recreation Centers, Libraries, Street Lights - **Question:** "Are high-crime areas underserved by community resources?" - **Operations:** Aggregate crime counts to census tracts or neighborhoods, count community resources per area, spatial correlation analysis - **Policy relevance:** Community investment, violence prevention strategies ---

#### Infrastructure & Services

**Option E: Polling Place Accessibility** - **Data:** Polling Places, SEPTA stops, Census tracts (elderly population, disability rates) - **Question:** "Are polling places accessible for elderly and disabled voters?" - **Operations:** Buffer polling places and transit stops, identify vulnerable populations, find areas lacking access - **Policy relevance:** Voting rights, election infrastructure, ADA compliance

------------------------------------------------------------------------

#### Health & Wellness

**Option F: Recreation & Population Health** - **Data:** Recreation Centers, Playgrounds, Parks, Census tracts (demographics) - **Question:** "Is lack of recreation access associated with vulnerable populations?" - **Operations:** Calculate recreation facilities per capita by neighborhood, buffer facilities for walking access, overlay with demographic indicators - **Policy relevance:** Public health investment, recreation programming, obesity prevention

------------------------------------------------------------------------

#### Emergency Services

**Option G: EMS Response Coverage** - **Data:** Fire Stations, EMS stations, Population density, High-rise buildings - **Question:** "Are population-dense areas adequately covered by emergency services?" - **Operations:** Create service area buffers (5-minute drive = \~2 miles), assess population coverage, identify gaps in high-density areas - **Policy relevance:** Emergency preparedness, station siting decisions

------------------------------------------------------------------------

#### Arts & Culture

**Option H: Cultural Asset Distribution** - **Data:** Public Art, Museums, Historic sites/markers, Neighborhoods - **Question:** "Do all neighborhoods have equitable access to cultural amenities?" - **Operations:** Count cultural assets per neighborhood, normalize by population, compare distribution across demographic groups - **Policy relevance:** Cultural equity, tourism, quality of life, neighborhood identity

------------------------------------------------------------------------

### Data Sources

**OpenDataPhilly:** https://opendataphilly.org/datasets/ - Most datasets available as GeoJSON, Shapefile, or CSV with coordinates - Always check the Metadata for a data dictionary of the fields.

**Additional Sources:** - **Pennsylvania Open Data:** https://data.pa.gov/ - **Census Bureau (via tidycensus):** Demographics, economic indicators, commute patterns - **TIGER/Line (via tigris):** Geographic boundaries

### Recommended Starting Points

**If you're feeling confident:** Choose an advanced challenge with multiple data layers. **If you are a beginner, choose something more manageable that helps you understand the basics**

**If you have a different idea:** Propose your own question! Just make sure: - You can access the spatial data - You can perform at least 2 spatial operations

### Your Analysis

**Your Task:**

1.  **Find and load additional data**
    -   Document your data source
    -   Check and standardize the CRS
    -   Provide basic summary statistics

```{r}
# Load your additional dataset




```

**Questions to answer:** - What dataset did you choose and why? - What is the data source and date? - How many features does it contain? - What CRS is it in? Did you need to transform it?

------------------------------------------------------------------------

2.  **Pose a research question**

Write a clear research statement that your analysis will answer.

**Examples:** - "Do vulnerable tracts have adequate public transit access to hospitals?" - "Are EMS stations appropriately located near vulnerable populations?" - "Do areas with low vehicle access have worse hospital access?"

------------------------------------------------------------------------

3.  **Conduct spatial analysis**

Use at least TWO spatial operations to answer your research question.

**Required operations (choose 2+):** - Buffers - Spatial joins - Spatial filtering with predicates - Distance calculations - Intersections or unions - Point-in-polygon aggregation

**Your Task:**

```{r}
# Your spatial analysis










```

**Analysis requirements:** - Clear code comments explaining each step - Appropriate CRS transformations - Summary statistics or counts - At least one map showing your findings - Brief interpretation of results (3-5 sentences)

**Your interpretation:**

\[Write your findings here\]

## Finally - A few comments about your incorporation of feedback!

Take a few moments to clean up your markdown document and then write a line or two or three about how you may have incorporated feedback that you recieved after your first assignment.

------------------------------------------------------------------------

## Submission Requirements

**What to submit:**

1.  **Rendered HTML document posted to your course portfolio** with all code, outputs, maps, and text
    -   Use `embed-resources: true` in YAML so it's a single file
    -   All code should run without errors
    -   All maps and charts should display correctly

**File naming:** `LastName_FirstName_Assignment2.html` and `LastName_FirstName_Assignment2.qmd`

------------------------------------------------------------------------