---
title: "Assignment 2: Spatial Analysis and Visualization"
subtitle: "Healthcare Access and Equity in Pennsylvania"
author: "Zhiyuan Zhao"
date: 2025/10/02
format: 
  html:
    code-fold: false
    toc: true
    toc-location: left
    theme: cosmo
    embed-resources: true
execute:
  warning: false
  message: false
---

## Assignment Overview

**Learning Objectives:** - Apply spatial operations to answer policy-relevant research questions - Integrate census demographic data with spatial analysis - Create publication-quality visualizations and maps - Work with spatial data from multiple sources - Communicate findings effectively for policy audiences

------------------------------------------------------------------------

## Part 1: Healthcare Access for Vulnerable Populations

### Research Question

**Which Pennsylvania counties have the highest proportion of vulnerable populations (elderly + low-income) living far from hospitals?**

Your analysis should identify counties that should be priorities for healthcare investment and policy intervention.

### Required Analysis Steps

Complete the following analysis, documenting each step with code and brief explanations:

#### Step 1: Data Collection (5 points)

Load the required spatial data: - Pennsylvania county boundaries - Pennsylvania hospitals (from lecture data) - Pennsylvania census tracts

**Your Task:**

```{r}
# Load required packages
library(sf)
library(tidyverse)
library(tidycensus)
library(tigris)
library(ggplot2)
library(knitr)
library(kableExtra)
library(units)
library(patchwork)
library(ggspatial)
```

```{r, include=FALSE}
census_api_key("a8081579173d9c4b9d21b9265f9c67c8708d6a02", install = TRUE, overwrite = TRUE)
```
```{r}
# Load spatial data
library(here)

pa_counties <- st_read(here::here("Assignments", "assignment_2", "data", "Pennsylvania_County_Boundaries.shp"))
hospitals   <- st_read(here::here("Assignments", "assignment_2", "data", "hospitals.geojson"))
# Pennsylvania census tracts
pa_tracts <- tracts("PA", year = 2022, cb = TRUE, progress = FALSE)

# Check that all data loaded correctly
cat("Number of hospitals:", nrow(hospitals), "\n")
cat("Number of census tracts", nrow(pa_tracts), "\n")
# CRS info
cat("Counties CRS:", st_crs(pa_counties)$input, "\n")
cat("Hospitals CRS:", st_crs(hospitals)$input, "\n")
cat("Census tracts CRS:", st_crs(pa_tracts)$input, "\n")

```

**Questions to answer:** 
1. How many hospitals are in your dataset? 
223
2. How many census tracts?
3445 
3. What coordinate reference system is each dataset in?
Counties CRS: WGS 84 / Pseudo-Mercator 
Hospitals CRS: WGS 84 
Census tracts CRS: NAD83 
------------------------------------------------------------------------

#### Step 2: Get Demographic Data

Use `tidycensus` to download tract-level demographic data for Pennsylvania.

**Required variables:** - Total population - Median household income - Population 65 years and over (you may need to sum multiple age categories)

**Your Task:**

```{r}
# Get demographic data from ACS
pa_demo <- get_acs(
  geography = "tract",
  state = "PA",
  variables = c(
    total_pop = "B01003_001",           
    median_income = "B19013_001",       
    pop_65_74 = "B01001_020",          
    pop_75_84 = "B01001_021",           
    pop_85p = "B01001_022",
    progress = FALSE         
  ),
  year = 2022,
  geometry = TRUE
)

# Pivot the data to wide format
pa_demo_wide <- pa_demo %>%
  select(GEOID, NAME, variable, estimate) %>%
  pivot_wider(
    names_from = variable, values_from = estimate,
    values_fill = 0
  ) %>%
  mutate(
    pop_65p = rowSums(across(c(pop_65_74, pop_75_84, pop_85p)), na.rm = TRUE),
    elderly_pct = 100 * pop_65p / total_pop
  )

# Join to tract boundaries
tracts_with_demo <- pa_demo_wide

# Check for missing data
cat("Missing income data:", sum(is.na(tracts_with_demo$median_income)), "\n")
cat("PA Median income: $", 
    median(tracts_with_demo$median_income, na.rm = TRUE), "\n")

```

**Questions to answer:** 
- What year of ACS data are you using? 
2022
- How many tracts have missing income data?
63
- What is the median income across all PA census tracts?
70188

------------------------------------------------------------------------

#### Step 3: Define Vulnerable Populations

Identify census tracts with vulnerable populations based on TWO criteria: 1. Low median household income (choose an appropriate threshold) 2. Significant elderly population (choose an appropriate threshold)

**Your Task:**

```{r}
# Filter for vulnerable tracts based on your criteria
# Define thresholds
income_threshold <- 50000   
elderly_threshold <- 10    

# Identify vulnerable tracts
vulnerable_tracts <- tracts_with_demo %>%
  filter(!is.na(median_income)) %>%
  mutate(
    low_income = median_income < income_threshold,
    sign_elderly = elderly_pct > elderly_threshold,
    vulnerable = low_income | sign_elderly
  )

# Summaries
summary_vals <- list(
  pa_median = median(tracts_with_demo$median_income, na.rm = TRUE),
  n_vul = sum(vulnerable_tracts$vulnerable),
  pct_vul = round(mean(vulnerable_tracts$vulnerable) * 100, 1),
  n_low = sum(vulnerable_tracts$low_income),
  n_old = sum(vulnerable_tracts$sign_elderly)
)

cat(glue::glue(
  "Income threshold: ${income_threshold} \n",
  "Elderly threshold: {elderly_threshold}%\n",
  "Vulnerable tracts: {summary_vals$n_vul} ({summary_vals$pct_vul}%)\n",
  "Low median household income: {summary_vals$n_low}\n",
  "Significant elderly population: {summary_vals$n_old}\n"
))



```

**Questions to answer:** 
- What income threshold did you choose and why? 
Income threshold chosen: $50,000.
This represents census tracts where the median household income is below roughly 70% of Pennsylvania’s overall median, marking areas that are relatively economically disadvantaged.

- What elderly population threshold did you choose and why? 
Elderly population threshold chosen: 10%.
Tracts where more than 10% of residents are aged 65 or older were considered to have a significant elderly population, a common demographic baseline in vulnerability analyses.

- How many tracts meet your vulnerability criteria? 
700 census tracts meet at least one of the two conditions 
- What percentage of PA census tracts are considered vulnerable by your definition?
20.7% 

------------------------------------------------------------------------

#### Step 4: Calculate Distance to Hospitals

For each vulnerable tract, calculate the distance to the nearest hospital.

**Your Task:**

```{r}
# Transform to appropriate projected CRS
# EPSG 2272 is optimized for southern Pennsylvania counties such as Philadelphia, Lancaster, Chester, and York
target_crs <- 2272

# Reproject all layers
pa_counties_p <- st_transform(pa_counties, target_crs)
hospitals_p   <- st_transform(hospitals,   target_crs)
vul_p         <- st_transform(vulnerable_tracts, target_crs)

# Calculate distance from each tract centroid to nearest hospital
# tract centroids
vul_centroids <- st_centroid(vul_p)

# nearest hospital index
nn_idx <- st_nearest_feature(vul_centroids, hospitals_p)

# distance in feet -> miles
dists_ft <- st_distance(vul_centroids, hospitals_p[nn_idx, ], by_element = TRUE)
dists_mi <- units::set_units(dists_ft, "mi") |> units::drop_units()

# attach results
vulnerable_tracts_dist <- vul_p |>
  mutate(
    nearest_hosp_idx = as.integer(nn_idx),
    dist_to_hosp_mi  = dists_mi
  )

# join hospital name if a name field exists
name_field <- intersect(c("NAME","Name","name","HOSP_NAME","HOSPITAL","HospName"), names(hospitals_p))[1]
if (!is.na(name_field)) {
  vulnerable_tracts_dist <- vulnerable_tracts_dist |>
    mutate(nearest_hosp_name = hospitals_p[[name_field]][nearest_hosp_idx])
}

# summary statistics
avg_dist <- mean(vulnerable_tracts_dist$dist_to_hosp_mi, na.rm = TRUE)
min_dist <- min(vulnerable_tracts_dist$dist_to_hosp_mi, na.rm = TRUE)
max_dist <- max(vulnerable_tracts_dist$dist_to_hosp_mi, na.rm = TRUE)
n_over15 <- sum(vulnerable_tracts_dist$dist_to_hosp_mi > 15, na.rm = TRUE)

cat(
  "Average distance to nearest hospital:", round(avg_dist, 2), "miles\n",
  "Minimum distance:", round(min_dist, 2), "miles\n",
  "Maximum distance:", round(max_dist, 2), "miles\n",
  "Number of tracts more than 15 miles from nearest hospital:", n_over15, "\n"
)



```

**Requirements:** - Use an appropriate projected coordinate system for Pennsylvania - Calculate distances in miles - Explain why you chose your projection

**Questions to answer:** 
- What is the average distance to the nearest hospital for vulnerable tracts? 
4.43 miles

- What is the maximum distance? 
32.84 miles

- How many vulnerable tracts are more than 15 miles from the nearest hospital?
127

------------------------------------------------------------------------

#### Step 5: Identify Underserved Areas

Define "underserved" as vulnerable tracts that are more than 15 miles from the nearest hospital.

**Your Task:**

```{r}
# Create underserved variable
# Definition: vulnerable tracts with distance > 15 miles

vulnerable_tracts_dist <- vulnerable_tracts_dist %>%
  mutate(underserved = dist_to_hosp_mi > 15)

# Counts and percentages
n_underserved  <- sum(vulnerable_tracts_dist$underserved, na.rm = TRUE)
n_with_dist    <- sum(!is.na(vulnerable_tracts_dist$dist_to_hosp_mi))
n_vulnerable   <- nrow(vulnerable_tracts_dist)
pct_underserved <- round(100 * n_underserved / n_with_dist, 1)

cat(
  "Underserved tracts:", n_underserved, "\n",
  "Percent underserved (of tracts with a distance):", pct_underserved, "%\n",
  "Vulnerable tracts total:", n_vulnerable, "\n",
  "With valid distance:", n_with_dist, "\n"
)



```

**Questions to answer:** 
- How many tracts are underserved? 
127

- What percentage of vulnerable tracts are underserved? 
3.8%

- Does this surprise you? Why or why not?
Yes, a bit less than I expected.
------------------------------------------------------------------------

#### Step 6: Aggregate to County Level

Use spatial joins and aggregation to calculate county-level statistics about vulnerable populations and hospital access.

**Your Task:**

```{r}
# find county name field in pa_counties_p
county_name_field <- intersect(
  c("NAME","NAME20","NAMELSAD","COUNTY","COUNTY_NAM","COUNTYNAME","County","county","NAME10"),
  names(pa_counties_p)
)[1]

if (is.na(county_name_field)) {
  stop("Cannot find a county name field in pa_counties_p. Inspect names(pa_counties_p).")
}

# keep only county name + geometry, and rename to COUNTY
counties_min <- pa_counties_p %>%
  dplyr::select(COUNTY = all_of(county_name_field), geometry)

# spatial join tracts -> counties
tracts_county <- st_join(vulnerable_tracts_dist, counties_min, join = st_intersects, left = FALSE)

# aggregate by COUNTY
county_summary <- tracts_county %>%
  st_drop_geometry() %>%
  dplyr::group_by(COUNTY) %>%
  dplyr::summarise(
    n_vulnerable        = dplyr::n(),
    n_underserved       = sum(underserved, na.rm = TRUE),
    pct_underserved     = round(100 * n_underserved / n_vulnerable, 1),
    avg_dist_mi         = round(mean(dist_to_hosp_mi, na.rm = TRUE), 2),
    total_vulnerable_pop= sum(total_pop, na.rm = TRUE)
  ) %>%
  dplyr::arrange(dplyr::desc(pct_underserved))

# table + tops
kable(county_summary, caption = "County-level Healthcare Accessibility Summary") %>%
  kable_styling(full_width = FALSE)

top5_pct <- county_summary %>%
  dplyr::slice_max(pct_underserved, n = 5) %>%
  dplyr::select(COUNTY, pct_underserved)

top5_pop <- county_summary %>%
  dplyr::slice_max(total_vulnerable_pop, n = 5) %>%
  dplyr::select(COUNTY, total_vulnerable_pop)

cat("Top 5 counties by % underserved:\n"); print(top5_pct)
cat("\nTop 5 counties by total vulnerable population:\n"); print(top5_pop)

```

**Required county-level statistics:** - Number of vulnerable tracts - Number of underserved tracts\
- Percentage of vulnerable tracts that are underserved - Average distance to nearest hospital for vulnerable tracts - Total vulnerable population

**Questions to answer:** 
- Which 5 counties have the highest percentage of underserved vulnerable tracts? 
JUNIATA, SULLIVAN, PIKE, SNYDER, CAMERON

- Which counties have the most vulnerable people living far from hospitals? 
PHILADELPHIA, ALLEGHENY, MONTGOMERY, BUCKS, CHESTER

- Are there any patterns in where underserved counties are located?
Yes. Underserved counties show clear geographic patterns:
	•	Spatial distribution: They are concentrated in rural, mountainous, and sparsely populated regions of Pennsylvania—especially in central and northern counties such as Juniata, Cameron, Sullivan, and Pike.
	•	Infrastructure gap: These areas have low hospital density and limited transportation access, leading to longer travel times to medical facilities.
	•	Urban contrast: Southeastern and southwestern metropolitan areas (Philadelphia, Allegheny, Montgomery, Bucks, Chester) have good hospital coverage and almost no tracts beyond 15 miles.

------------------------------------------------------------------------

#### Step 7: Create Summary Table

Create a professional table showing the top 10 priority counties for healthcare investment.

**Your Task:**

```{r}
# Create and format priority counties table
# Priority rule: sort by % underserved (desc), then avg distance (desc),
# then # underserved tracts (desc), then vulnerable population (desc)

priority_tbl <- county_summary %>%
  arrange(desc(pct_underserved), desc(avg_dist_mi), desc(n_underserved), desc(total_vulnerable_pop)) %>%
  slice_head(n = 10) %>%
  transmute(
    County = COUNTY,
    `Vulnerable tracts`   = scales::comma(n_vulnerable),
    `Underserved tracts`  = scales::comma(n_underserved),
    `% underserved`       = scales::number(pct_underserved, accuracy = 0.1),
    `Avg distance (mi)`   = scales::number(avg_dist_mi, accuracy = 0.01),
    `Vulnerable population` = scales::comma(total_vulnerable_pop)
  )

kable(
  priority_tbl,
  caption = "Top 10 Priority Counties for Healthcare Investment (sorted by % underserved, distance, underserved count, population)"
) %>%
  kable_styling(full_width = FALSE)



```

**Requirements:** - Use `knitr::kable()` or similar for formatting - Include descriptive column names - Format numbers appropriately (commas for population, percentages, etc.) - Add an informative caption - Sort by priority (you decide the metric)

------------------------------------------------------------------------

## Part 2: Comprehensive Visualization

Using the skills from Week 3 (Data Visualization), create publication-quality maps and charts.

### Map 1: County-Level Choropleth

Create a choropleth map showing healthcare access challenges at the county level.

**Your Task:**

```{r}
library(ggplot2)
library(scales)
library(dplyr)

# join county-level stats to county geometries
counties_map <- pa_counties_p %>%
  left_join(county_summary, by = c("COUNTY_NAM" = "COUNTY"))

# choropleth filled by % underserved; hospitals as points
ggplot() +
  geom_sf(data = counties_map, aes(fill = pct_underserved), color = NA) +
  geom_sf(data = hospitals_p, size = 0.6, alpha = 0.7) +
  scale_fill_viridis_c(
    name = "% underserved tracts",
    labels = label_percent(accuracy = 0.1, scale = 1),
    na.value = "grey90"
  ) +
  labs(
    title = "Healthcare Access Gaps by County in Pennsylvania",
    subtitle = "% of vulnerable tracts > 15 miles from nearest hospital",
    caption = "Sources: ACS 2022 (tidycensus), PA hospitals, county boundaries"
  ) +
  theme_void() +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold")
  )
```

**Requirements:** - Fill counties by percentage of vulnerable tracts that are underserved - Include hospital locations as points - Use an appropriate color scheme - Include clear title, subtitle, and caption - Use `theme_void()` or similar clean theme - Add a legend with formatted labels

------------------------------------------------------------------------

### Map 2: Detailed Vulnerability Map

Create a map highlighting underserved vulnerable tracts.

**Your Task:**

```{r}
# Create detailed tract-level map
# Data prep
tracts_map <- vulnerable_tracts_dist
underserved_tracts <- tracts_map %>% filter(underserved)

ggplot() +
  # context: county boundaries
  geom_sf(data = pa_counties_p, fill = NA, color = "grey50", linewidth = 0.2) +
  # all vulnerable tracts (muted)
  geom_sf(data = tracts_map, fill = "grey85", color = NA, alpha = 0.6) +
  # underserved vulnerable tracts (highlight)
  geom_sf(data = underserved_tracts, aes(fill = dist_to_hosp_mi), color = NA) +
  # hospitals
  geom_sf(data = hospitals_p, shape = 21, fill = "black", color = "white", size = 1) +
  scale_fill_viridis_c(
    name = "Distance to hospital (mi)",
    option = "C",
    limits = c(15, max(underserved_tracts$dist_to_hosp_mi, na.rm = TRUE)),
    oob = scales::squish
  ) +
  labs(
    title = "Underserved Vulnerable Tracts in Pennsylvania",
    subtitle = "Tracts with vulnerable populations located > 15 miles from the nearest hospital",
    caption = "Sources: ACS 2022 (tidycensus), PA hospitals, county boundaries"
  ) +
  theme_void() +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold")
  )



```

**Requirements:** - Show underserved vulnerable tracts in a contrasting color - Include county boundaries for context - Show hospital locations - Use appropriate visual hierarchy (what should stand out?) - Include informative title and subtitle

------------------------------------------------------------------------

### Chart: Distribution Analysis

Create a visualization showing the distribution of distances to hospitals for vulnerable populations.

**Your Task:**

```{r}
# Create distribution visualization
# Histogram of distances (vulnerable tracts)
p_hist <- ggplot(vulnerable_tracts_dist, aes(x = dist_to_hosp_mi)) +
  geom_histogram(bins = 40) +
  geom_vline(xintercept = 15, linetype = "dashed") +
  labs(
    title = "Distance to Nearest Hospital (Vulnerable Tracts)",
    x = "Distance (miles)", y = "Count",
    caption = "Dashed line at 15 miles threshold"
  ) +
  theme_minimal()

# Top-10 counties by underserved tracts (requires county_summary from Step 6)
top10_underserved <- county_summary %>%
  slice_max(n_underserved, n = 10) %>%
  mutate(COUNTY = reorder(COUNTY, n_underserved))

p_bar <- ggplot(top10_underserved, aes(x = COUNTY, y = n_underserved)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top 10 Counties by Number of Underserved Tracts",
    x = NULL, y = "Underserved tracts (count)"
  ) +
  theme_minimal()

# Scatter: distance vs. population, highlighting underserved tracts
p_scatter <- ggplot(vulnerable_tracts_dist,
                    aes(x = total_pop, y = dist_to_hosp_mi, alpha = underserved)) +
  geom_point() +
  scale_x_continuous(labels = comma) +
  scale_alpha_manual(values = c(`TRUE` = 1, `FALSE` = 0.3), guide = "none") +
  labs(
    title = "Distance vs. Tract Population (Vulnerable Tracts)",
    x = "Tract total population", y = "Distance to nearest hospital (miles)"
  ) +
  theme_minimal()

# Display
(p_hist | p_scatter) / p_bar

# Brief interpretation
cat("Interpretation:",
    "Most vulnerable tracts lie within 15 miles, but a right tail exceeds the threshold.",
    "Underserved tracts concentrate in a small set of counties.",
    "Higher-population tracts are not necessarily farther from hospitals; gaps are mainly geographic.\n")



```

**Suggested chart types:** - Histogram or density plot of distances - Box plot comparing distances across regions - Bar chart of underserved tracts by county - Scatter plot of distance vs. vulnerable population size

**Requirements:** - Clear axes labels with units - Appropriate title - Professional formatting - Brief interpretation (1-2 sentences as a caption or in text)

------------------------------------------------------------------------

## Part 3: Bring Your Own Data Analysis

Choose your own additional spatial dataset and conduct a supplementary analysis.

### Your Analysis

**Your Task:**

1.  **Find and load additional data**
    -   Document your data source
    -   Check and standardize the CRS
    -   Provide basic summary statistics

```{r}
# Config
crs_proj <- 2272          # Pennsylvania South State Plane (US feet)
buf_ft   <- 2640          # 0.5 mile ≈ 2640 ft

# Load schools 
schools <- st_read(here("Assignments", "assignment_2", "data","Schools.geojson"), quiet = TRUE) %>%
  st_transform(crs_proj)

# Get Philadelphia tracts with children population (ACS 2022)
# B09001_001 = total population under 18
acs_children <- get_acs(
  geography = "tract",
  variables = "B09001_001",
  state = "PA",
  county = "Philadelphia",
  year = 2022,
  geometry = TRUE
) %>%
  st_transform(crs_proj) %>%
  rename(child_pop = estimate) %>%
  select(GEOID, NAME, child_pop, geometry) %>%
  mutate(
    tract_name = str_replace(NAME, ", Philadelphia County, Pennsylvania", "")
  )

# Build Philadelphia boundary from tracts
phila_boundary <- acs_children %>% st_union() %>% st_as_sf()

# Data overview
cat("\n===== DATA OVERVIEW =====\n")

cat("\n-- Schools --\n")
print(schools)
cat("\nCRS:\n"); print(st_crs(schools))
cat("\nField summary:\n"); glimpse(schools)
cat("\nCoordinate summary:\n"); print(summary(st_coordinates(schools)))

cat("\n-- Philadelphia tracts (ACS 2022) --\n")
print(acs_children)
cat("\nCRS:\n"); print(st_crs(acs_children))
cat("\nChild population summary:\n"); print(summary(acs_children$child_pop))

# Quick visualization check
ggplot() +
  geom_sf(data = acs_children, fill = "grey95", color = NA) +
  geom_sf(data = schools, color = "blue", size = 0.6) +
  labs(title = "Spatial extent check – Schools and Philadelphia tracts") +
  theme_minimal()


```

**Questions to answer:**
- What dataset did you choose and why? 
Philadelphia public schools and 2022 ACS census tracts with children population (under 18 years old). The school dataset defines the physical locations of educational facilities, while the tract-level ACS data provides demographic context to measure how many children live within and outside the school service areas.

- What is the data source and date? - How many features does it contain? 
The school locations were downloaded from OpenDataPhilly (“Philadelphia Schools”).
The children population data were retrieved from the U.S. Census Bureau’s American Community Survey (ACS) 5-Year Estimates, 2022 release via the tidycensus R package.
The OpenDataPhilly schools layer contains about 300 school points within Philadelphia.
The ACS dataset includes 384 census tracts for the City of Philadelphia, each with a child_pop attribute (population under 18).

- What CRS is it in? Did you need to transform it?
Both datasets were standardized to EPSG:2272 — Pennsylvania South State Plane (US feet) to ensure accurate distance and area calculations.
The original school data used a geographic coordinate system (latitude/longitude), so it was transformed to EPSG:2272 before creating 0.5-mile buffers. The ACS data were downloaded with geometry and then also projected to EPSG:2272 for consistency.


------------------------------------------------------------------------

2.  **Pose a research question**

Write a clear research statement that your analysis will answer.

**Examples:** - "Do vulnerable tracts have adequate public transit access to hospitals?" - "Are EMS stations appropriately located near vulnerable populations?" - "Do areas with low vehicle access have worse hospital access?"

Do areas in Philadelphia with higher concentrations of children have adequate spatial access to schools within a 0.5-mile walking distance?
------------------------------------------------------------------------

3.  **Conduct spatial analysis**

Use at least TWO spatial operations to answer your research question.

**Required operations (choose 2+):** - Buffers - Spatial joins - Spatial filtering with predicates - Distance calculations - Intersections or unions - Point-in-polygon aggregation

**Your Task:**

```{r}
# Your spatial analysis
# Build 0.5-mile service buffer of schools
school_buf <- schools %>%
  st_buffer(buf_ft) %>%      # service area
  st_union() %>%             # dissolve overlaps
  st_as_sf()

# Tract-level coverage and areal-weighted children counts
area_total   <- st_area(acs_children)
overlap_idx  <- st_intersects(acs_children, school_buf)
has_overlap  <- lengths(overlap_idx) > 0
area_covered <- units::set_units(0, "ft^2")
area_covered[has_overlap] <- st_area(st_intersection(acs_children[has_overlap, ], school_buf))

tracts_cov <- acs_children %>%
  mutate(
    area_total   = area_total,
    area_covered = area_covered,
    pct_covered  = pmin(1, as.numeric(area_covered / area_total)),
    children_in_service = child_pop * pct_covered,
    children_in_gap     = child_pop * (1 - pct_covered),
    access_flag = case_when(
      pct_covered >= 0.90 ~ "Well covered (≥90%)",
      pct_covered >= 0.50 ~ "Partially covered (50–90%)",
      TRUE               ~ "Gap (<50%)"
    )
  )


# Citywide summaries
city_summary <- tracts_cov %>%
  st_drop_geometry() %>%
  summarise(
    tracts              = n(),
    children_total      = sum(child_pop, na.rm = TRUE),
    children_served     = sum(children_in_service, na.rm = TRUE),
    children_gap        = sum(children_in_gap, na.rm = TRUE),
    pct_children_served = round(100 * children_served / children_total, 1),
    pct_children_gap    = round(100 * children_gap / children_total, 1)
  )
cat("\n===== CITYWIDE SUMMARY =====\n")
print(city_summary)

band_summary <- tracts_cov %>%
  st_drop_geometry() %>%
  group_by(access_flag) %>%
  summarise(
    n_tracts     = n(),
    children_sum = sum(child_pop, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(match(access_flag, c("Well covered (≥90%)","Partially covered (50–90%)","Gap (<50%)")))
cat("\n===== COVERAGE BANDS =====\n")
print(band_summary)

top_gap_tracts <- tracts_cov %>%
  st_drop_geometry() %>%
  transmute(
    tract_name,
    GEOID,
    child_pop,
    pct_covered = round(pct_covered, 3),
    children_in_gap = round(children_in_gap)
  ) %>%
  arrange(desc(children_in_gap)) %>%
  slice_head(n = 10)
cat("\n===== TOP 10 GAP TRACTS =====\n")
print(top_gap_tracts)

# Maps
# Coverage percent choropleth
ggplot() +
  geom_sf(data = tracts_cov, aes(fill = pct_covered), color = NA) +
  geom_sf(data = phila_boundary, fill = NA, color = "grey40", linewidth = 0.2) +
  scale_fill_viridis_c(name = "Coverage by area", labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "School Service Coverage by Tract — Philadelphia",
    subtitle = "0.5 mile buffer around schools; tract coverage by area overlap",
    caption = "Schools: OpenDataPhilly; Children: ACS 2022 (B09001_001); CRS: EPSG 2272"
  ) +
  theme_minimal()

# Children outside service areas (areal-weighted)
ggplot() +
  geom_sf(data = tracts_cov, aes(fill = children_in_gap), color = NA) +
  geom_sf(data = phila_boundary, fill = NA, color = "grey40", linewidth = 0.2) +
  scale_fill_viridis_c(name = "Children outside 0.5 mi", labels = scales::comma) +
  labs(
    title = "Children Outside 0.5-mile School Service Areas",
    subtitle = "Areal-weighted estimate per tract",
    caption = "ACS 2022 (B09001_001); CRS: EPSG 2272"
  ) +
  theme_minimal()

```

**Analysis requirements:** - Clear code comments explaining each step - Appropriate CRS transformations - Summary statistics or counts - At least one map showing your findings - Brief interpretation of results (3-5 sentences)

**Your interpretation:**

Philadelphia tracts show high access overall: 95.1% of children live within 0.5 miles of a school; 4.8% are outside service areas. Coverage by tract: 322 well covered (≥90%), 57 partially covered (50–90%), 29 gaps (<50%), totaling 293,209, 41,085, and 8,291 children respectively. The largest unmet-need tracts by children outside service areas include Census Tracts 348.02, 347.01, 363.03, 361, 373, 358, 357.02, 365.02, 346, 337.01. Maps confirm a small number of edge tracts drive most gaps; prioritizing these for new sites or access improvements would yield the highest impact.

## Finally - A few comments about your incorporation of feedback!

I used include=FALSE to hide my API, removed the empty [], deleted some overly long instructions and improved the variable names to make the code easier to read.

------------------------------------------------------------------------

